---
title: "IgScan Installation"
author: 
  name: Ian Márquez-López
  email: iamarquez@recerca.clinic.cat
  affiliation: Institut d’Investigacions Biomèdiques August Pi i Sunyer (FRCB-IDIBAPS)
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IgScan install}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Installing from GitHub (Recommended)

To install the development version of **IgScan** (v.1.0.0) from GitHub:

```{r, eval=FALSE}
install.packages("devtools")
library(devtools)

devtools::install_github("IanMarquezIDIBAPS/IgScan", build_vignettes = T, dependencies = T)
```

## Installing from Local Source

If you have downloaded or cloned the IgScan repository locally:

#### Using R

```{r, eval=FALSE}
devtools::install("path/to/IgScan", build_vignettes = T, dependencies = T)
```

#### From the terminal

```{r, eval=FALSE}
R CMD INSTALL path/to/IgScan/
```

Make sure you have a working R development environment (Rtools on Windows, Xcode on macOS, or build tools on Linux).

## Accessing Vignettes

After installation, you can explore the package vignettes (detailed tutorials) by running...

```{r, eval=FALSE}
browseVignettes("IgScan")
```

... or access them through the IgScan website.
