---
title: "IgScan Installation"
author: 
  name: Ian Márquez-López
  email: iamarquez@recerca.clinic.cat
  affiliation: Institut d’Investigacions Biomèdiques August Pi i Sunyer (FRCB-IDIBAPS)
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IgScan install}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Installing from GitHub (Recommended)

To install the development version of **IgScan** (v.0.1.0) from GitHub:

```{r, eval=FALSE}

## First, install Bioconductor dependencies
if (!requireNamespace("BiocManager", quietly = TRUE)){
  install.packages("BiocManager") 
}

BiocManager::install(
  c(
    "SingleCellExperiment",
    "pwalign",
    "Biostrings",
    "IRanges",
    "GenomicAlignments",
    "treeio", 
    "ggtree"
  ),
  ask = FALSE,
  update = FALSE
)

## Install IgScan using the remotes package
install.packages("remotes")
remotes::install_github("nadeulab/IgScan", build_vignettes = T, dependencies = T)
```

## Installing from Local Source

If you have downloaded or cloned the IgScan repository locally:

#### Using R

```{r, eval=FALSE}
devtools::install("path/to/IgScan", build_vignettes = T, dependencies = T)
```

#### From the terminal

```{r, eval=FALSE}
R CMD INSTALL path/to/IgScan/
```

Make sure you have a working R development environment (Rtools on Windows, Xcode on macOS, or build tools on Linux).

## System requirements: IgBlast on Apple Silicon (M1/M2/M3/M4/M5)

IgScan relies on the external `igblastn` executable from NCBI IgBLAST. If you are running macOS on an Apple Silicon Mac (M1, M2, M3, M4 or M5), be aware that the standard IgBlast binaries distributed by NCBI are compiled for Intel processors (`x86_64`), not for Apple's native `arm64` architecture.

If `igblastn` is not natively compatible with your CPU, you will see an error such as:

```
zsh: bad CPU type in executable
```

### How to check if this affects you

Run the following commands in your terminal:

```bash
# Check your Mac's architecture
uname -m

# Check the architecture the igblastn binary was built for
file $(which igblastn)
```

If `uname -m` returns `arm64` but `file` reports `x86_64`, your system needs Rosetta 2 to run IgBlast.

### Solution: install Rosetta 2

Rosetta 2 is Apple's translation layer that allows Intel-built executables to run on Apple Silicon Macs. Install it by running:

```bash
softwareupdate --install-rosetta --agree-to-license
```

Once installed, `igblastn` (and any other Intel-only dependency required by IgScan) will run normally, without any further configuration.

```{r, echo=FALSE, eval=FALSE}
# Note for developers: this chunk is not evaluated.
# Rosetta 2 only needs to be installed once per machine.
```

## Accessing Vignettes

After installation, you can explore the package vignettes (detailed tutorials) by running...

```{r, eval=FALSE}
browseVignettes("IgScan")
```

... or access them through the IgScan website.
